---
output: rmarkdown::github_document
---

# Army 17, Air Force 14

# Can we predict the number of yards Army gained each play on offense?
This code performs analysis of play-by-play data from the Army - Air Force 2018 football game.  Intended to demonstrate regression models in introductory statistics courses.

The file "PBP - 2018 - Week 10.csv" contains play-by-play data for every college football game up to Week 10.  The data is available from user BlueSCar at https://www.reddit.com/r/CFBAnalysis/comments/6wz5ww/2017_play_by_play_data/. 

## Load libraries and data
```{r, message = FALSE}
library(tidyverse)
plays <- read_csv(file = "PBP - 2018 - Week 10.csv")
View(plays)
```

## Data Cleaning
```{r, message = FALSE}
# First, we will filter out other games.
plays <- plays %>% filter(homeTeam == "Army")

# Next, filter out punts, kickoffs, penalties, etc. 
unique(plays$type)
plays_RushAndPass <- plays %>% 
  filter(type %in% c("Rush","Pass Reception",
                     "Rushing Touchdown",
                     "Pass Incompletion")) 
```

## Data Analysis
```{r, message=FALSE}
#Yards by quarter
plays_RushAndPass %>% 
  ggplot(aes(x = factor(quarter), y = yardsGained, color = offenseTeam)) +
  geom_boxplot() + ggtitle("Yards gained from scrimmage each play by quarter")

#Yards by down
plays_RushAndPass %>% ggplot(aes(x = factor(down), y = yardsGained, color = offenseTeam)) +
  geom_boxplot() + ggtitle("Yards gained from scrimmage each play by down")

plays_RushAndPass %>% group_by(quarter, offenseTeam) %>% 
  summarise(averageYards = mean(yardsGained),
            totalYards = sum(yardsGained), 
            n = n())
```

## Question: Does Army gain more yards when they have farther to go for a first down?

In this case, our response variable is yards gained and the predictor is number of yards to go for a first down.

### Some plots
```{r, message = FALSE}
# filter out when Army is on defense
plays_Army <- plays_RushAndPass %>% filter(offenseTeam == "Army") 

plays_Army %>% ggplot(aes(x = distance, y = yardsGained)) + 
  geom_point() + geom_smooth(method = "lm") + ggtitle("yards gained vs. distance to go")
```

Some questions to consider:

* Is a linear model appropriate?
* Are there any outliers?
* What does a slope of zero mean?

### Fit a linear model 
```{r}
model_distance <- lm(yardsGained ~ distance, 
                 data = plays_Army)
summary(model_distance)
```

Some questions:

* How do we interpret the distance coefficient? Intercept?
* Is this result statistically significant? What does that mean?
* What does the residual standard error mean?
* What does this model tell us?

## Perhaps, we can learn more about yards gained by including other information.

Let's see what happens when we include what down it is.

### Some plots
```{r}
plays_Army %>% ggplot(aes(x = factor(down), y = yardsGained)) +
  geom_boxplot()
```

### Fit a model including down
```{r}
model_distance_down <- lm(yardsGained ~ distance + factor(down), 
                          data = plays_Army)
summary(model_distance_down)
```

Some questions:

* How do we interpret the coefficients in this model?
* How would the model change if we did not put factor() around downs in the linear model?
* The distance coefficient changed a lot from the simple model.  What are some explanations for that?
* What happened to the residual standard error? 

